---
title: "Simulation of Dissolution Profiles with Predefined Target $f_2$"
author: "Zhengguo XU"
date: "`r format(Sys.Date())`"
output: 
  rmarkdown::html_vignette:
    keep_md: true
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 4.5
  highlight: "tango"
bibliography: ref.bib
notes-after-punctuation: false
link-citations: yes
csl: ref.csl
vignette: >
  %\VignetteIndexEntry{Simulation of Dissolution Profiles with Predefined Target $f_2$}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup0, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```

```{r setup, echo=FALSE}
library(bootf2)
```

## Introduction
Sometimes you need to simulate dissolution profiles with known $f_2$ values
to study its properties. It is relatively easy to do by trial and error, 
but that takes times. This is where the function `sim.dp.byf2` comes into play.

The basic principle is as follows:

1. For any given mean dissolution profile, the function will fit a suitable
   mathematical model and obtain model parameters. 
2. Based on the model parameters from the preliminary fitting, 
   the function will generate another set of parameters and simulate 
   a new dissolution profile.
3. Compare the new dissolution profile to the target profile by calculating
   $f_2$.
4. If the obtained $f_2$ equal to the target $f_2$, the function will output 
   the model parameters and the simulated profile. If not, step 2 and 3 will 
   be repeated. 

A few notes on the workflow above:

- If the output of function `sim.dp()` is available, obviously no fitting is
  necessary since the output of `sim.dp()` contained all modelling information.
- There are two ways to set the target $f_2$ value:
    - If target $f_2$ is a *single value*, optimization algorithm will be used 
      and the simulated dissolution profile will have $f_2$ equal to the
      target $f_2$ when compared to the given profile (within the numeric
      precision defined by the tolerance).
    - If target $f_2$ is vector of two numbers, such as c(lower, upper), 
      representing the lower and upper limit of target $f_2$ value, then
      dissolution will be obtained by random searching and the calculated 
      $f_2$ will be within the range of lower and upper. 
    - For example, to have profile with target $f_2$ of 55, you can set 
      target $f_2$ to `c(54.95, 55.04)`. Consider that $f_2$ should usually 
      be reported without decimal, the precision is enough in practical use.
      You can also use `c(54.99, 55.01)` to have higher precisin. However, 
      the higher precision, the longer the time it will take the function 
      to run. With setting such as  `c(54.999, 55.001)`, it is likely the
      program will hang up. In such case, provide single value to use
      optimization algorithm. 


## Usage
The complete list of arguments of the function is as follows:

```{r simdpbyf2-code0, eval=FALSE}
sim.dp.byf2(tp, dp, sim.dp.out, target.f2, seed,
            regulation = c("EMA", "FDA", "WHO"),
            model = c("Weibull", "first-order"), digits = 2L,
            min.points = 3L, both.TR.85 = FALSE, max.disso = 105,
            model.par.cv = 50, fix.fmax.cv = 0, random.factor = 3,
            sim.target = c("ref.pop", "ref.median", "ref.mean"),
            time.unit = c("min", "h"), message = TRUE, plot = TRUE)
```

### Notes on function arguments
1. The input should be either `tp` and `dp` for the time points and target
   dissolution profile to which the newly simulated profiles will be compared,
   or `sim.dp.out`, the output of function `sim.dp()` if available. 
    - If `sim.dp.out` is provided together with `tp` and `dp`, the latter
      two will be ignored.
1. Option `target.f2 = 55` or `target.f2 = c(54.99, 55.01)`. See the 
   explanation in previous section.
1. Option `model.par.cv` is used for the random generation of model parameters
   by $P_i = P_\mu \cdot e^{N\left(0,\, \sigma^2\right)}$, where 
   $\sigma = \mathrm{CV}/100$. The default value works most of the time. In 
   rare cases when the function does not return any value or gives error
   message indicating that no parameters can be find, it might be helpful to
   change it to higher value. It is only applicable when `target.f2` is
   provided as lower and upper limit.
1. Option `fix.fmax.cv` is similar to `model.par.cv` above but just for the 
   parameter `fmax` since it is usually fixed at 100. If this parameter
   should also be varied, set it to non-zero value such 3 or 5.
1. Option `random.factor` is also used for the generation of model parameters
   but what make it different from `model.par.cv` and `fix.fmax.cv` is that 
   it is only used when `target.f2` is provided as a singe value. Similarly, 
   the default value should work most of the time so only change it
   when the function does not work properly.
1. To use $f_2$ method, one of the condition is that there should be at
   least 3 time points, which is controlled by option `min.points = 3`.
   Therefore, if the provided dissolution `dp` is a very fast release
   profile and there is not enough time points before 85% dissolution,
   sometime it is impossible to find a new profile. For example, if the 
   profile dissolve more than 85% at the second time point, $f_2$ method 
   cannot be used. In such case, the function will return error message.
   You can set the `min.points` to a smaller value such as 2. 
1. Option `sim.target` is a character strings indicating to which target
   dissolution profile should the newly simulated be used to compare by 
   calculating f2. This is only applicable when `sim.dp.out` is provided
   because the output of `sim.dp()` contains the population profile and 
   the descriptive statistics (e.g., mean and median) of the simulated 
   individual profiles. If only `tp` and `dp` are provided, then `dp` is
   considered as the population profile. See examples below.
1. See vignette [Calculating Similarity Factor $f_2$](calcf2.md) and function
   manual by `help("sim.dp.byf2")` for details of the rest options. 

## Examples

### With output from function `sim.dp()`

Simulate a reference profile. 
```{r simdpbyf2-01}
# time points
tp <- c(5, 10, 15, 20, 30, 45, 60)

# model.par for reference
par.r <- list(fmax = 100, fmax.cv = 3, mdt = 15, mdt.cv = 14, 
              tlag = 0, tlag.cv = 0, beta = 1.5, beta.cv = 8)

# simulate reference data
dref <- sim.dp(tp, model.par = par.r, seed = 100)
```
Now find another (test) profile that has predefined $f_2$ of 50.

```{r simdpbyf2-f2-50a}
df2_50_a <- sim.dp.byf2(sim.dp.out = dref, target.f2 = 50, seed = 123)
```
We can check how close is the calculated $f_2$ to the target $f_2$.
```{r simdpbyf2-f2-50a1}
format(df2_50_a$model.par$f2 - 50, scientific = FALSE)
```


Obviously, change seed number will usually produce a different result.
```{r simdpbyf2-f2-50-b2}
df2_50_b <- sim.dp.byf2(sim.dp.out = dref, target.f2 = 50, seed = 234)

# precision 
format(df2_50_b$model.par$f2 - 50, scientific = FALSE)
```

```{r simdpbyf2-f2-50-3}
df2_50_c <- sim.dp.byf2(sim.dp.out = dref, target.f2 = c(49.99, 50.01),
                        seed = 456)

# check to see that this is less precise, but still enough for practical use
format(df2_50_c$model.par$f2 - 50, scientific = FALSE)
```

### With `tp` and `dp`
The input can be just a vector of time points `tp` and mean profiles `dp`
```{r simdpbyf2-f2-55a}
dp <- c(17, 42, 63, 78, 94, 99, 100)

df2_55a <- sim.dp.byf2(tp, dp, target.f2 = 55, seed = 100)
# check precision
format(df2_55a$model.par$f2 - 55, scientific = FALSE)
```

Similarly, target $f_2$ can be a range.

```{r simdpbyf2-f2-55b}
df2_55b <- sim.dp.byf2(tp, dp, target.f2 = c(54.95, 55.04), seed = 100)

# check precision
format(df2_55b$model.par$f2 - 55, scientific = FALSE)
```

### More advanced examples

The output of the function contains a list of 2 components, 
a data frame of model parameters `model.par` and a data frame of mean
dissolution `sim.disso`. The former is actually more useful since it can
be used as input for function `sim.dp()` to simulate individual profiles. 
For example, data `dref` contains individual dissolution profiles (see
vignette [Simulation of Dissolution Profiles](sim.dp.md) for details.),
we can simulate many individual profiles that will have predefined $f_2$
values by finding the suitable model parameters first, then use 
`sim.dp()` in the second step. Or use the simulated mean profile as input
of `dp` for `sim.dp()` in the second step.

For example, to simulate individual profile with $f_2 = 54$ when compared 
to the mean of the 12 units of reference profile, we can do the following:

Simulate a profiles that has $f_2 = 54$ compared to the reference mean 
profile.
 
```{r simdpbyf2-adv1}
# get model parameters
par.t54 <- sim.dp.byf2(sim.dp.out = dref, target.f2 = 54, 
                       sim.target = "ref.mean", seed = 1111)

# check precision
format(par.t54$model.par$f2 - 54, scientific = FALSE)
```


Then use the newly simulated profile as input for the function `sim.dp()`.

```{r simdpbyf2-adv2}
# use mean profile as input for sim.dp()
dp2 <- par.t54$sim.disso$test[-1]
print(dp2)

# simulate test profiles
dtest <- sim.dp(tp, dp = dp2, seed = 2222)
```


Compare the profiles of test and reference.

```{r simdpbyf2-adv3}
# compare to reference profile
t_vs_r <- calcf2(dtest$sim.disso, dref$sim.disso)

# check precision
format(t_vs_r[1] - 54, scientific = FALSE)
```
