---
title: "Simulation of Dissolution Profiles with Predefined Target $f_2$"
author: "Zhengguo XU"
date: "`r format(Sys.Date())`"
output: 
  rmarkdown::html_vignette:
    keep_md: true
    toc: false
    toc_depth: 3
    fig_width: 7
    fig_height: 4.5
  highlight: "tango"
bibliography: ref.bib
notes-after-punctuation: false
link-citations: yes
csl: ref.csl
vignette: >
  %\VignetteIndexEntry{Simulation of Dissolution Profiles with Predefined Target $f_2$}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup0, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```

```{r setup, echo=FALSE}
library(bootf2)
```

## Introduction
Sometimes you need to simulate dissolution profiles with known $f_2$ values
to study its properties. It is relatively easy to do by trial and error, but
that takes times. This is where the function `sim.dp.byf2` comes into play.

The main principle of the function is as follows:

1. For any given mean dissolution profile `dp`, fit a suitable mathematical
   model and obtain model parameters.
    - No precise fitting is required since those parameters will be served as
      *initial value* for further model fitting.
    - If `sim.dp.out`, the output of the function `sim.dp()`, is available, 
      no initial fitting is necessary as model parameters can be read directly
      from the output, unless multivariate normal distribution approach was 
      used in `sim.dp()`. In such case, initial model fitting will be done.
1. Find a suitable model parameters and simulate a new dissolution profile,
   comparing the new profile to the provided reference profile `dp` by
   calculating $f_2$. If the the obtained $f_2$ is equal to, or within the 
   lower and upper limit of, the `target.f2`, then the function will output 
   the obtained model parameters and the new profile.

There are two approaches used to find the suitable model parameters:

- If `target.f2` is a single value, optimization algorithm will be used and the
  newly simulated dissolution profile will have $f_2$ equal to `target.f2` when
  compared to `dp` (within numeric precision defined by the tolerance).

- If `target.f2` is a vector of two numbers representing the lower and upper
  limit of target $f_2$ value, such as `target.f2 = c(lower, upper)`, then 
  dissolution will be obtained by random searching and the calculated $f_2$ 
  will be within the range of lower and upper.

For example, you can set `target.f2 = c(54.95, 55.04)` to have target $f_2$ of
55. Since $f_2$ should be normally reported without decimal, in practice, this
precision is enough. You might be able to do with `c(54.99, 55.01)` if you
really need more precision. However, the narrower the range, the longer it 
takes the function to run. With narrow range such as `c(54.999, 55.001)`, it is
likely the program will fail. In such case, provide single value to use
optimization algorithm.

Arguments `model.par.cv`, `fix.fmax.cv`, and `random.factor` are certain
numeric values used to better control the random generation of model parameters.
The default values should work in most scenarios. Those values should be changed
only when the function failed to return any value. See more details below.

The data frame `sim.summary` in `sim.dp.out`, the output of function `sim.dp()`,
contains `dp`, the *population* profile, and `sim.mean` and `sim.median`, the
mean and median profiles calculated with `n.units` of simulated individual
profiles. All these three profiles could be used as the target profile that the
newly simulated profile will be compare to. Argument `sim.target` defines which
of the three will be used: `ref.pop`, `ref.mean`, and `ref.median` correspond 
to `dp`, `sim.mean` and `sim.median`, respectively.

The output of the function is a list of 2 components: a data frame of model
parameters and a data frame of mean dissolution profile generated using the
said model parameters. The output can be used as input to function `sim.dp()` 
to further simulate multiple individual profiles.

## Usage
The complete list of arguments of the function is as follows:

```{r simdpbyf2-code0, eval=FALSE}
sim.dp.byf2(tp, dp, target.f2, seed = NULL, min.points = 3L,
            regulation = c("EMA", "FDA", "WHO", "Canada", "ANVISA"),
            model = c("Weibull", "first-order"), digits = 2L,
            max.disso = 100, message = FALSE, both.TR.85 = FALSE,
            time.unit = c("min", "h"), plot = TRUE, sim.dp.out,
            sim.target = c("ref.pop", "ref.median", "ref.mean"),
            model.par.cv = 50, fix.fmax.cv = 0, random.factor = 3)
```

### Notes on function arguments
1. The input should be either `tp` and `dp` for the time points and target
   dissolution profile to which the newly simulated profiles will be compared,
   or `sim.dp.out`, the output of function `sim.dp()` if available. 
    - If `sim.dp.out` is provided together with `tp` and `dp`, the latter
      two will be ignored.
1. Option `model.par.cv` is used for the random generation of model parameters
   by $P_i = P_\mu \cdot e^{N\left(0,\, \sigma^2\right)}$, where 
   $\sigma = \mathrm{CV}/100$. The default value works most of the time. In 
   rare cases when the function does not return any value or gives error
   message indicating that no parameters can be find, it might be helpful to
   change it to higher value. It is only applicable when `target.f2` is
   provided as lower and upper limit.
1. Option `fix.fmax.cv` is similar to `model.par.cv` above but just for the 
   parameter `fmax` since it is usually fixed at 100. If this parameter
   should also be varied, set it to non-zero value such 3 or 5.
1. Option `random.factor` is also used for the generation of model parameters
   but what make it different from `model.par.cv` and `fix.fmax.cv` is that 
   it is only used when `target.f2` is provided as a singe value. Similarly, 
   the default value should work most of the time so only change it
   when the function does not work properly.
1. To use $f_2$ method, one of the condition is that there should be at least 3
   time points, which is controlled by option `min.points = 3`. Therefore, if
   the provided dissolution `dp` is a very fast release profile and there is
   not enough time points before 85% dissolution, sometime it is impossible to
   find a new profile. For example, if the profile dissolve more than 85% at 
   the second time point, $f_2$ method cannot be used. In such case, the
   function will return error message. You can set the `min.points` to a 
   smaller value such as 2. 
1. Option `sim.target` is a character strings indicating to which target
   dissolution profile should the newly simulated be used to compare by 
   calculating f2. This is only applicable when `sim.dp.out` is provided
   because the output of `sim.dp()` contains the population profile and the
   descriptive statistics (e.g., mean and median) of the simulated individual
   profiles. If only `tp` and `dp` are provided, then `dp` is considered as the
   population profile. See examples below.
1. See vignette *Calculating Similarity Factor $f_2$* and function manual by
   `help("sim.dp.byf2")` for details of the rest options. 

## Examples

### With output from function `sim.dp()`

Simulate a reference profile. 
```{r simdpbyf2-01}
# time points
tp <- c(5, 10, 15, 20, 30, 45, 60)

# model.par for reference
par.r <- list(fmax = 100, fmax.cv = 3, mdt = 15, mdt.cv = 14, 
              tlag = 0, tlag.cv = 0, beta = 1.5, beta.cv = 8)

# simulate reference data
dref <- sim.dp(tp, model.par = par.r, seed = 100)
```
Now find another (test) profile that has predefined $f_2$ of 50.

```{r simdpbyf2-f2-50a}
df2_50_a <- sim.dp.byf2(sim.dp.out = dref, target.f2 = 50, seed = 123,
                        message = TRUE)
```
We can check how close is the calculated $f_2$ to the target $f_2$.
```{r simdpbyf2-f2-50a1}
format(df2_50_a$model.par$f2 - 50, scientific = FALSE)
```


Obviously, change seed number will usually produce a different result.
```{r simdpbyf2-f2-50-b2}
df2_50_b <- sim.dp.byf2(sim.dp.out = dref, target.f2 = 50, seed = 234,
                        message = TRUE)

# precision 
format(df2_50_b$model.par$f2 - 50, scientific = FALSE)
```

```{r simdpbyf2-f2-50-3}
df2_50_c <- sim.dp.byf2(sim.dp.out = dref, target.f2 = c(49.99, 50.01),
                        seed = 456, message = TRUE)

# check to see that this is less precise, but still enough for practical use
format(df2_50_c$model.par$f2 - 50, scientific = FALSE)
```

### With `tp` and `dp`
The input can be just a vector of time points `tp` and mean profiles `dp`
```{r simdpbyf2-f2-55a}
dp <- c(17, 42, 63, 78, 94, 99, 100)

df2_55a <- sim.dp.byf2(tp, dp, target.f2 = 55, seed = 100,
                       message = TRUE)
# check precision
format(df2_55a$model.par$f2 - 55, scientific = FALSE)
```

Similarly, target $f_2$ can be a range.

```{r simdpbyf2-f2-55b}
df2_55b <- sim.dp.byf2(tp, dp, target.f2 = c(54.95, 55.04), seed = 100,
                       message = TRUE)

# check precision
format(df2_55b$model.par$f2 - 55, scientific = FALSE)
```
